---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# FacileArchs4Data

This package is a work in progress and aims to implement the Facile API over
the [ARCHS4][] (v2) datasets.

Currently the package doesn't provide much functionality except for easy
wrappers to construct bioconductor expression containers (DGEList,
SummarizedExperiment, etc.) out of whatever subset of samples and genes
that you like.

[arcsh4]: https://archs4.org/

## Installation

You'll need to download the archs4 hdf5 files yourself form their
[download page][archs4dl]. I put them in versioned directories somewhere on my
system, like:

```
~/workspace/data/archs4/v2.6/human_gene_v2.6.h5
~/workspace/data/archs4/v2.6/mouse_gene_v2.6.h5
```

Then install this package from github:

```r
pak::pak("facilebio/FacileArchs4Data")
```

[archs4dl]: https://archs4.org/download

## Usage

Currently you should use the `a4 <- Archs4Client$new("mouse")` (or `"human"`) R6
constructor to get a thin wrapper around the data object. Then, subset the
`a4$samples` tibble to include the the samples you want to analyze. Finally,
convert the sample-frame it to a `DGEList` or `SummarizedExperiment` to move on
with your life -- see the **Usage** section for details.

To make your life easy, store the "equally versioned" mouse and human
datasets in a single directory (`$ARCHS4_DATADIR`) with the value of `species`
in the filename, and set a global R option (`archs4.data_dir`) to
`$ARCHS4_DATADIR`.

You can put new/old h5 files in different directories, then load the dataset by
specifying simply the species. See the **Installation** section for more
details.

Note that when you load the hdf5 file, a subset of columns will be loaded
for all of the samples by default. This is the (likely few second delay) you
experience when calling the `$new()` constructor for the first time.


```{r, message=FALSE, warning=FALSE}
library(FacileArchs4Data)
library(dplyr)

# NOTE: You can set this in your ~/.Rprofile file to make your life easier
options(archs4.data_dir = "~/workspace/data/archs4/v2.6")
a4h <- Archs4Client$new("human")
```

## Finding samples / expreiments of interest

Some of the funtions in this package are fashioned after the archs4py package,
so there is a `$search()` function made available to you that enables you to
search the metadata of each sample based on some of the decault columns.

Let's say you were interested in searching for "parksinsons" related samples:

```{r}
samples.park <- a4h$search("parkinson")
```

This will return a tibble of samples where the word `"parkinson"` appears
in one of hte meta data fields:

```{r}
head(samples.park)
```

The sample metadata tibble is also easily available in the `$samples` tibble,
and you can search for rows of interest "as you are used to".

## Expression Analysis

Let's say you already knew you were interested in using the `"GSE40710"` dataset
(aSyn polyA-RNAseq in PD and unaffected cortical brain samples) to explore the
gene expression differencs between healthy and parkinson's brain transcriptomes.

We can easily extract the samples in this dataset like so:

```{r}
samples <- filter(a4h$samples, series_id == "GSE40710")
```

Your next challenge will be to identify which samples belong to healthy vs
affected individuals. The data in archs4 matches what is provided in the GEO
database, and there are unfortunately no standards as far as data/sample
annotations go, so each individual experiment will provide you with its own
set of challenges.

Often, the sample-level metadata you will be most interested will be in the
`"source_name_ch1"` and `"title"` columns:

```{r}
set.seed(0xBEEF)
samples |> 
  sample_n(6) |> 
  select(sample, title, source_name_ch1)
```

We can see that the parkinson's vs control status is in the `"title"` column.
Let's great a "group" column with the affected status, and then use the
`biocbox` function in this package to retrieve the associated expression data
for these samples and turn all of it into a DGEList for downstream analysis.

```{r}
y <- samples |> 
  mutate(
    # extract control/parksinson & remove the title column
    group = sub(" .*", "", title),
    # the "'" to make `group` a well behaved R variable
    group = sub("'", "", group)
  ) |> 
  biocbox(class = "DGEList")
```

Now you are ready to analyze `y` using your favorite edgeR/limma workflow.

## FacileData Integration

If you have the FacileBiocData and FacileAnalysisShine package installed, you
can then use `y` for interactive analysis.

```{r, message=FALSE, warning=FALSE}
library(FacileBiocData)
library(FacileAnalysisShine)
yf <- facilitate(y)
```

### Principal Components Analysis

```{r}
ypca <- fpca(yf)
viz(ypca, color_aes = "group")
```

### Differential Expressoin Analysis

```{r}
ydge <- yf |> 
  flm_def(covariate = "group", numer = "Parkinsons", denom = "Control") |> 
  fdge(method = "voom")
viz(ydge, type = "volcano", ntop = 2000, webgl = FALSE)
```

Refer to the [FacileAnalysis][] documentation for more analysis fun.

## License

Please note that the use of archs4 data is subject to its own agreement /
licensing restrictions. Be sure to follow those.

The code in this package is released under the [Apache 2 License][apache2].

## Acknowledgements

Thanks to Alexander Lachmann and The Ma’ayan Laboratory for the development
and continued updates to the [ARCHS4 project][archs4].

[FacileAnalysis]: https://facilebio.github.io/FacileAnalysis/
[facilebio]: https://facile.bio/
[apache2]: https://www.apache.org/licenses/LICENSE-2.0
