---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# FacileArchs4DataSet

This package implements the Facile API over the ARCHS4 (v2) datasets, and provides some fundamental code required to build the children dataset repositories for the human and mouse [HDF5 files][archs4dl] files provided by the ARCHS4 project.

These [ARCHS4 dataset][archs4] repositories can then be loaded into a `FacileArchs4DataSet` object, which is a subclass of a `FacileDataSet` and subsequently used by the [facilebio][] ecosystem.

## Installation

To build the species-specifc dataset packages, users will need to download the (v2) gene level HDF5 files for the species of interest (mouse or human) from the [ARCHS4 project][archs4dl]
and run the `build` command:

```{r, eval = FALSE}
# pak::pkg_install("facilebio/FacileArchs4DataSet")
FacileArchs4DataSet::build(
  h5 = "/path/to/archs4_gene_human_v2.1.2.h5",
  outdir = "/path/to/repository/dir",
  species = "human")
```

This will create a `/path/to/repository/dir/human` directory that can be used to instantiate a `FacileArcsh4DataSet` that can be used in the facilebio ecosystem.

## Data Exploration

Once the data packge is successfully built, you can use it for exploratory analysis, the facile way:

```{r, eval = FALSE}
library(FacileAnalysis)

a4h <- FacileArchs4DataSet::load("/path/to/repository/dir/human")
a4h |>
  filter_samples(dataset == "GSE89189") |>
  fdgeGadget()
```

Refer to the [FacileAnalysis][] documentation for details on how to easily
perform diferential expression analysis, GSEA, etc.

## License

In order for this package to be useful, the end user must download the HDF5 files provifed by the [ARCHS4 project][archs4dl]. Use of the ARCHS4 data is subject to its own [terms of use][archs4tos], and it is the end-user's responsibility to ensure that they are using it in a compliant manner.

The code in this package that facilitates the query and retrieval of the data provided is released under [The Apache 2 License][apache2].

## Acknowledgements

Thanks toAlexander Lachmann and The Ma’ayan Laboratory for the development andcontinued updates to the [ARCHS4 project][archs4].

[archs4]: https://maayanlab.cloud/archs4/index.html
[archs4dl]: https://maayanlab.cloud/archs4/download.html
[archs4tos]: https://maayanlab.cloud/archs4/help.html
[FacileAnalysis]: https://facilebio.github.io/FacileAnalysis/
[facilebio]: https://facile.bio/
[apache2]: https://www.apache.org/licenses/LICENSE-2.0