---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# FacileArchs4DataSet

This package is a work in progress and aims to implement the Facile API over
the ARCHS4 (v2) datasets. It provides some fundamental code required to build
the children dataset repositories for the human and mouse [HDF5 files][archs4dl]
files provided by the ARCHS4 project.

These [ARCHS4 dataset][archs4] repositories can then be loaded into a
`FacileArchs4DataSet` object, which is a subclass of a `FacileDataSet` and subsequently used by the [facilebio][] ecosystem.

This package is currently written against [v2.2][archs4dl] of the human and
mouse file format.

## State of Affairs

Currently you should use the `a4 <- Archs4Client$new("mouse")` (or `"human"`) R6
constructor to get a thin wrapper around the data object. Then, subset the
`a4$samples` tibble to include the the samples you want to analyze. Finally,
convert the sample-frame it to a `DGEList` or `SummarizedExperiment` to move on
with your life -- see the **Usage** section for details.

To make your life easy, store the "equally versioned" mouse and human
datasets in a single directory (`$ARCHS4_DATADIR`) with the value of `species`
in the filename, and set a global R option (`archs4.data_dir`) to
`$ARCHS4_DATADIR`.

You can put new/old h5 files in different directories, then load the dataset by
specifying simply the species. See the **Installation** section for more
details.

## Usage

```{r}
library(FacileArchs4DataSet)
# NOTE: Set this in your ~/.Rprofile file to make your life easier
options(archs4.data_dir = "~/workspace/data/archs4/v2.2")
a4h <- Archs4Client$new("human")
a4h$samples |> head()
```

## Installation

Create a directory (`$ARCHS4_DATADIR`) on your local filesystem and download the
mouse and human gene expression count matrices from the
[ARCHS4 Download page][archs4dl]. Let's assume
`ARCHS4_DATA_DIR='~/data/archs4/v2.2'`

This directory should look like:

```bash
$ ls -l $ARCHS4_DATA_DIR # ls -l ~/data/archs4/v2.2
-rw-r--r--@ 1 steve  staff    35G Jan 22 14:25 human_gene_v2.2.h5
-rw-r--r--@ 1 steve  staff    28G Feb  5 16:13 mouse_gene_v2.2.h5
```

In your ~/.Rprofile, set the `archs4.data_dir` option to `$ARCHS4_DATADIR`:

```bash
$ cat ~/.Rprofile
options(
  repos = c(cloud = "https://cloud.r-project.org"),
  archs4.data_dir = "~/data/archs4/v2.2")
```

Now install this package from GitHub.


```r
remotes::install_github("facilebio/FacileArchs4DataSet")
```

## Usage

Find a human dataset of interest and convert it into a `SummarizedExperiment`:

```{r}
library(FacileArchs4DataSet)
a4h <- Archs4Client$new("human")
# Let's get a DGEList of some iPSC microglia data from GSE186301
y <- a4h$samples |> 
  dplyr::filter(series_id == "GSE186301") |> 
  biocbox(class = "DGEList")
y$samples |> 
  dplyr::select(dataset, sample_id, source_name_ch1) |> 
  dplyr::sample_n(5)
```

NOTE: The rest of these instructions was from v1 of this software. There were
some good ideas there, so leaving it for reference but they don't work.

## Installation

To build the species-specifc dataset packages, users will need to download the (v2) gene level HDF5 files for the species of interest (mouse or human) from the [ARCHS4 project][archs4dl]
and run the `build` command:

```{r, eval = FALSE}
# pak::pkg_install("facilebio/FacileArchs4DataSet")
FacileArchs4DataSet::build(
  h5 = "/path/to/archs4_gene_human_v2.1.2.h5",
  outdir = "/path/to/repository/dir",
  species = "human")
```

This will create a `/path/to/repository/dir/human` directory that can be used to instantiate a `FacileArcsh4DataSet` that can be used in the facilebio ecosystem.

## Data Exploration

Once the data packge is successfully built, you can use it for exploratory analysis, the facile way:

```{r, eval = FALSE}
library(FacileAnalysis)

a4h <- FacileArchs4DataSet::load("/path/to/repository/dir/human")
a4h |>
  filter_samples(dataset == "GSE89189") |>
  fdgeGadget()
```

Refer to the [FacileAnalysis][] documentation for details on how to easily
perform diferential expression analysis, GSEA, etc.

## License

In order for this package to be useful, the end user must download the HDF5 files provifed by the [ARCHS4 project][archs4dl]. Use of the ARCHS4 data is subject to its own [terms of use][archs4tos], and it is the end-user's responsibility to ensure that they are using it in a compliant manner.

The code in this package that facilitates the query and retrieval of the data provided is released under [The Apache 2 License][apache2].

## Acknowledgements

Thanks toAlexander Lachmann and The Ma’ayan Laboratory for the development andcontinued updates to the [ARCHS4 project][archs4].

[archs4]: https://maayanlab.cloud/archs4/index.html
[archs4dl]: https://maayanlab.cloud/archs4/download.html
[archs4tos]: https://maayanlab.cloud/archs4/help.html
[FacileAnalysis]: https://facilebio.github.io/FacileAnalysis/
[facilebio]: https://facile.bio/
[apache2]: https://www.apache.org/licenses/LICENSE-2.0
