---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  message = FALSE,
  error = FALSE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

# FacileArchs4DataSet

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/facilebio/FacileArchs4DataSet.svg?branch=master)](https://travis-ci.org/facilebio/FacileArchs4DataSet)
[![Codecov test coverage](https://codecov.io/gh/facilebio/FacileArchs4DataSet/branch/master/graph/badge.svg)](https://codecov.io/gh/facilebio/FacileArchs4DataSet?branch=master)
<!-- badges: end -->

[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Lifecycle: Experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

This package enables the use of the mouse and human [ARCHS4 datasets][archs4]
within the [facilebio][] ecosystem by defining a `FacileDataSet` subclass
(`FacileArchs4DataSet`) that uses the [HDF5 files][archs4dl] provided by the
ARCHS4 project as the assay data source, as opposed to the FacileDataSet-native
way of storing these data on disk.

The `FacileArchs4DataSet` class therefore provides the facile data API over
these data, thereby enabling its use in the [facilebio][] ecosystem.

## Installation

```{r, eval = FALSE}
# remotes::install("facilebio/FacileArchs4DataSet")
library(FacileArchs4DataSet)
```

1. Users need to download the both the gene and transcript level HDF5 files for
   the species of interest (mouse or human) from the [ARCHS4 project][archs4dl]
   into a new/empty folder, ie. 
   `ARCHS4DATADIR` = "/archs4dir/with/human/hdf5files".

2. Run the `FacileArchs4DataSet::build(ARCHS4DATADIR, species = "human")`

This will take a few seconds to build the appropriate sample and gene level
database. After which, the user can perform a differential expression analysis
over the data from the `"GSE89189"` study like so:

```{r, eval = FALSE}
library(FacileAnalysis)
a4.mouse <- FacileArchs4DataSet(ARCHS4DATADIR)
a4.mouse %>% 
  filter_samples(dataset == "GSE89189") %>% 
  fdgeGadget()
```

Refer to the [FacileAnalysis][] documentation for details on how to easily
perform diferential expression analysis, GSEA, etc.

## License

There is not data provided within this package, and the code within is released
under [The Apache 2 License][apache2].

In order for this package to be useful, however, the end user must download
the HDF5 files provifed by the [ARCHS4 project][archs4dl]. Use of the ARCHS4
data is subject to its own [terms of use][archs4tos], and it is the end-user's
responsibility to ensure that they are using it in a compliant manner.


## Acknowledgements

* Thanks Alexander Lachmann and The Ma’ayan Laboratory for the development and
  continued updates to the [ARCHS4 project][archs4].

[archs4]: https://amp.pharm.mssm.edu/archs4/index.html
[archs4dl]: https://amp.pharm.mssm.edu/archs4/download.html
[archs4tos]: https://amp.pharm.mssm.edu/archs4/help.html
[FacileAnalysis]: https://facilebio.github.io/FacileAnalysis/
[facilebio]: https://facile.bio/
